Interpreting Random Forest Predictions for Firearm Identification Using LIME

### 2019-01-16 journal/applying LIME to bullet data journal.Rmd
worked more on updating code to implement trees and started reorganizing journal

---
title: "Applying LIME to Bullet Data Journal"
author: "Katherine Goode"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE, fig.pos = "center")

# Load libraries
library(tidyverse)
library(bulletr)
library(randomForest)
library(lime)
library(cowplot)
library(purrr)
library(future)
library(furrr)

@@

## 12/5/18 - 12/14/18

I decided to start a journal this week to keep track of the work I am doing in regards to applying LIME to the Hamby bullet data. I have been trying to decide which input values to use for LIME in the forensic examiners' paper, and I have decided that I would like to simplify matters even more by only focusing on a few input options. I have considered a handful of ways of answering this question (accuracy within one implementation of lime, accuracy across 10 implementations of lime, etc.), but I would like to reserve some of these ways for a future more technical paper. However, I would like a place to store my work to use for future purposes, and it would be nice to have a place to work where I do not feel the need to optimize code. I am hoping that this will be a place where I can write drafts of code and create first versions of graphics. Then I can clean up and organize the code when I write the paper.

### Previous Work

#### Feature Plots of Training and Testing Data Filled by `samesource`

@@

#### Correlation Plots of Training and Testing Data

@@

#### Visualizations of the LIME Explanations

@@

#### Draft of LIME Procedure from the `lime` R Package

@@

### Work on Determining Input Values for LIME

@@

#### Assessing the Accuracy of the LIME Results

@@

##### Complex versus Simple Model Predictions

@@

```{r}
hamby224_lime_comparisons %>%
  mutate(quantile_bins = fct_recode(factor(quantile_bins), "equal bins" = "FALSE",
                                    "quantile bins" = "TRUE")) %>%
  filter(!is.na(rfscore), set == "Set 1", bin_continuous == TRUE) %>%
  ggplot(aes(x = rfscore, y = model_prediction)) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid(quantile_bins ~ nbins) + 
  theme_bw() + 
  stat_smooth(method = "lm", se = FALSE, size = 0.5) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

```{r}
hamby224_lime_comparisons %>%
  mutate(quantile_bins = fct_recode(factor(quantile_bins), "equal bins" = "FALSE",
                                    "quantile bins" = "TRUE")) %>%
  filter(!is.na(rfscore), set == "Set 1", bin_continuous == TRUE) %>%
  ggplot(aes(x = rfscore, y = abs(diff))) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid(quantile_bins ~ nbins) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

```{r}
hamby224_lime_comparisons %>%
  filter(quantile_bins == FALSE, nbins == "3", set == "Set 1") %>%
  select(rf_features, diff, -case) %>%
  gather(key = feature, value = feature_value, rf_features) %>%
  select(feature, feature_value, diff) %>%
  ggplot(aes(x = feature_value, y = diff)) + 
  geom_point() + 
  facet_wrap(~ feature, scales = "free") + 
  geom_hline(yintercept = 0, color = "blue")
```

##### Comparing Input Values by MSE and $R^2$

@@

```{r}
hamby224_lime_results <- hamby224_lime_comparisons %>%
  filter(!is.na(rfscore)) %>%
  group_by(bin_continuous, quantile_bins, nbins, use_density, set) %>%
  summarise(mse = (sum(diff^2)) / length(diff),
            ave_r2 = mean(model_r2)) %>%
  arrange(set) %>%
  ungroup() %>%
  mutate(nbins = as.numeric(as.character(nbins))) %>%
  mutate(situation = ifelse(bin_continuous == TRUE & quantile_bins == TRUE, 
                            sprintf("%.0f quantile bins", nbins),
                            ifelse(bin_continuous == TRUE & quantile_bins == FALSE, 
                                   sprintf("%.0f equally spaced bins", nbins),
                                   ifelse(bin_continuous == FALSE & use_density == TRUE, 
                                          "kernel density", "normal approximation")))) %>%
  mutate(situation = fct_relevel(situation, 
                                 "2 quantile bins", 
                                 "3 quantile bins",
                                 "4 quantile bins",
                                 "5 quantile bins",
                                 "6 quantile bins")) %>%
  mutate(bin_situation = ifelse(bin_continuous == TRUE & quantile_bins == TRUE,
                                "quantile bins",
                                ifelse(bin_continuous == TRUE & quantile_bins == FALSE,
                                       "equally spaced bins", "other")))
```

@@

```{r}
hamby224_lime_results %>%
  ggplot(aes(x = situation, y = mse, color = bin_situation)) +
  geom_point() + 
  facet_grid( ~ set) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom")
```

@@

```{r}
hamby224_lime_results %>%
  ggplot(aes(x = situation, y = ave_r2, color = bin_situation)) +
  geom_point() + 
  facet_grid( ~ set) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

#### Assessing the Variability of the LIME Results

@@

#### Deciding on LIME Input Settings

@@

#### Notes on meeting with Heike

@@

## 12/15/18 - 1/13/19

## 1/14/19 - ?

### Goals

Goals for spring semester:

- implement ideas for better binning methods
- write draft of firearm examiner's paper with everything but the methods and results sections
- find an idea that leads to better LIME explanations
- once the better LIME binning method has been selected, write the methods and results sections
- make plans for the statistical paper

Goals for week 1:

- set up meeting time with Heike
- get tree method up and running
- start to work on evaluation of the results of the new method
- start looking into the weights
- work on introduction and data section of firearm examiners' paper

### Visualizations of Tree Based Bins Results

```{r}
# I need to fix the situations to new options 
hamby224_test_explain %>%
  mutate(situation = ifelse(bin_continuous == TRUE & quantile_bins == TRUE, 
                            sprintf("%.0f quantile bins", nbins),
                            ifelse(bin_continuous == TRUE & quantile_bins == FALSE, 
                                   sprintf("%.0f equally spaced bins", nbins),
                                   ifelse(bin_continuous == FALSE & use_density == TRUE, 
                                          "kernel density", "normal approximation"))),
         diff = rfscore - model_prediction) %>%
  mutate(situation = fct_relevel(situation, 
                                 "2 quantile bins", 
                                 "3 quantile bins",
                                 "4 quantile bins",
                                 "5 quantile bins",
                                 "6 quantile bins")) %>%
  mutate(bin_situation = ifelse(bin_continuous == TRUE & quantile_bins == TRUE,
                                "quantile bins",
                                ifelse(bin_continuous == TRUE & quantile_bins == FALSE,
                                       "equally spaced bins", "other"))) %>%
  filter(set == "Set 1") %>%
  droplevels() %>%
  na.omit() %>%
  select(situation, case, feature, feature_weight) %>%
  ggplot(aes(x = case, y = situation, fill = feature)) + 
  geom_tile()
```

>>>>

---
title: "Applying LIME to Bullet Data Research Journal"
author: "Katherine Goode"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, 
                      fig.pos = "center")
```

# Overview

I decided to start a journal to keep track of the work I am doing in regards to applying LIME to the Hamby bullet data. I have been trying to decide which input values to use for LIME in the forensic examiners' paper, and I have decided that I would like to simplify matters even more by only focusing on a few input options. I have considered a handful of ways of answering this question (accuracy within one implementation of lime, accuracy across 10 implementations of lime, etc.), but I would like to reserve some of these ways for a future more technical paper. However, I would like a place to store my work to use for future purposes, and it would be nice to have a place to work where I do not feel the need to optimize code. I am hoping that this will be a place where I can write drafts of code and create first versions of graphics. Then I can clean up and organize the code when I write the paper.

```{r}
# Load libraries
library(tidyverse)
library(bulletr)
library(randomForest)
library(lime)
library(cowplot)
library(purrr)
library(future)
library(furrr)

@@

# Plots of Features

## Distributions by `samesource`

@@

## Correlations

@@

# Visualizations of Explanations

@@

# Draft of LIME Procedure

@@

# Tree Based Bins

## Functions

I wrote the function `mylime`, which adds an option in the `lime` function from the LIME package to use Heike's function `treebink` to obtain tree based bins. The output from `mylime` can then be input into the `explain` function to obtain explanations. I was able to rerun all of the code for creating the `hamby224_test_explain` dataset with the explanations from the tree based bins included for 2 to 6 bins. I've been working on redoing code from earlier in this journal to work for the added binning method.

There seems to be a problem with `treebink` with certain variables with high enough bins. The function returns more than the desired number of bins. Heike is going to look into this.

# Assessing Lime Explanations

@@

## Accuracy

@@

### Complex versus Simple Model Predictions

@@

```{r}
hamby224_lime_comparisons %>%
  filter(!is.na(rfscore), set == "Set 1", bin_continuous == TRUE) %>%
  ggplot(aes(x = rfscore, y = model_prediction)) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid(bin_method ~ nbins) + 
  theme_bw() + 
  stat_smooth(method = "lm", se = FALSE, size = 0.5) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

```{r}
hamby224_lime_comparisons %>%
  filter(!is.na(rfscore), set == "Set 1", bin_continuous == TRUE) %>%
  ggplot(aes(x = rfscore, y = abs(diff))) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid(bin_method ~ nbins) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

```{r}
hamby224_lime_comparisons %>%
  filter(quantile_bins == FALSE, bin_method == "equally_spaced", nbins == "3", set == "Set 1") %>%
  select(rf_features, diff, -case) %>%
  gather(key = feature, value = feature_value, rf_features) %>%
  select(feature, feature_value, diff) %>%
  ggplot(aes(x = feature_value, y = diff)) + 
  geom_point() + 
  facet_wrap(~ feature, scales = "free") + 
  geom_hline(yintercept = 0, color = "blue")
```

### Comparing Input Values by MSE and $R^2$

@@

```{r}
hamby224_lime_results <- hamby224_lime_comparisons %>%
  filter(!is.na(rfscore)) %>%
  group_by(situation, bin_situation, bin_continuous, quantile_bins, nbins, use_density, 
           bin_method, set) %>%
  summarise(mse = (sum(diff^2)) / length(diff),
            ave_r2 = mean(model_r2)) %>%
  arrange(set) %>%
  ungroup()
```

@@

```{r}
hamby224_lime_results %>%
  ggplot(aes(x = situation, y = mse, color = bin_situation)) +
  geom_point() + 
  facet_grid( ~ set) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom") + 
  labs(x = "Situation", y = "MSE", color = "")
```

@@

```{r}
hamby224_lime_results %>%
  ggplot(aes(x = situation, y = ave_r2, color = bin_situation)) +
  geom_point() + 
  facet_grid( ~ set) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") + 
  labs(x = "Situation", y = "Average R2", color = "")
```

## Variability

@@

## Input Rankings

@@

## Heike's Suggestions

@@

## Results from One Implementation

```{r}
# Selects only the top features for the cases where binning was used
top_features <- hamby224_test_explain %>%
  filter(bin_situation != "other") %>%
  select(set, situation, bin_situation, case, feature, feature_weight) %>%
  mutate(feature_weight_abs = abs(feature_weight)) %>%
  arrange(situation, case, desc(feature_weight_abs)) %>%
  group_by(situation, case) %>%
  slice(1)

# Plots the top feature for each case for each bin situation
ggplot(top_features, aes(x = situation, y = case, fill = feature)) + 
  geom_tile() + 
  facet_grid(set ~ bin_situation, scales = "free") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_brewer(palette = "Blues")

# Plots the proportions of number of top features chosen across number of bins for a case
# within a binning method
top_features %>%
  group_by(set, case, bin_situation) %>%
  summarise(nfeatures = length(levels(factor(feature)))) %>%
  ggplot(aes(x = bin_situation, fill = factor(nfeatures))) + 
  geom_bar(position = "stack") + 
  facet_grid(set ~ .) + 
  theme_bw() +
  scale_fill_brewer(palette = "Blues")
```

## Assessing Variability

### 2019-01-16 paper/Interpreting Random Forest Predictions for Firearm Identification Using LIME.Rmd
worked more on updating code to implement trees and started reorganizing journal

```{r setup, include = FALSE}
# Set knitr options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE, fig.pos = "center")

@@

  # Reorder the variables of feature_desc and feature_bin for plotting purposes
  hamby224_test_explain <- hamby224_test_explain %>%
    mutate(feature_desc = reorder(feature_desc, strictly_less),
           feature_desc = reorder(feature_desc, feature_number),
           feature_desc = reorder(feature_desc, as.numeric(feature))) %>%
    select(bin_continuous:bin_method, case:feature_desc, feature_bin:strictly_less, 
           data, prediction)

@@

  # Create a data frame with the interesting information relating to the different
  # evaluations of lime and compute the difference and mean between the rf and rr
  # model predictions
  hamby224_lime_comparisons <- hamby224_test_explain %>%
    select(-data, -prediction) %>%
    group_by(case, bin_continuous, quantile_bins, nbins, use_density, bin_method) %>%
    slice(1) %>%
    ungroup() %>%
    select(bin_continuous, quantile_bins, nbins, use_density, 
           set, case, rf_features, rfscore, model_prediction, model_r2) %>%
    mutate(diff = rfscore - model_prediction,
           mean = (rfscore + model_prediction) / 2)

>>>>

```{r setup, include = FALSE}
# Set knitr options
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, 
                      fig.pos = "center")

@@

  # Reorder the variables of feature_desc and feature_bin for plotting purposes and
  # create new variables of situation and bin_situation
  hamby224_test_explain <- hamby224_test_explain %>%
    mutate(feature_desc = reorder(feature_desc, strictly_less),
           feature_desc = reorder(feature_desc, feature_number),
           feature_desc = reorder(feature_desc, as.numeric(feature))) %>%
    mutate(nbins = as.numeric(as.character(nbins)),
           situation = ifelse(bin_continuous == TRUE & bin_method == "quantile_bins", 
                              sprintf("%.0f quantile bins", nbins),
                              ifelse(bin_continuous == TRUE & bin_method == "equally_spaced",
                                     sprintf("%.0f equally spaced bins", nbins),
                                     ifelse(bin_continuous == TRUE & bin_method == "tree",
                                            sprintf("%.0f tree based bins", nbins),
                                            ifelse(bin_continuous == FALSE & use_density == TRUE, 
                                                   "kernel density", 
                                                   "normal approximation")))) %>%
             fct_relevel("2 quantile bins", "3 quantile bins", "4 quantile bins",
                         "5 quantile bins", "6 quantile bins", "2 equally spaced bins",
                         "3 equally spaced bins", "4 equally spaced bins",
                         "5 equally spaced bins", "6 equally spaced bins")) %>%
    mutate(bin_situation = ifelse(bin_method == "quantile_bins" & bin_continuous == TRUE,
                                "quantile bins",
                                ifelse(bin_method == "equally_spaced" & bin_continuous == TRUE,
                                       "equally spaced bins", 
                                       ifelse(bin_method == "tree" & bin_continuous == TRUE,
                                              "tree based bins", "other")))) %>%
    mutate(bin_situation = factor(bin_situation)) %>%
    select(situation, bin_situation, bin_continuous:bin_method, case:feature_desc,
           feature_bin:strictly_less, data, prediction)

@@

  # Create a data frame with the interesting information relating to the different
  # evaluations of lime and compute the difference and mean between the rf and rr
  # model predictions
  hamby224_lime_comparisons <- hamby224_test_explain %>%
    select(-data, -prediction) %>%
    group_by(case, bin_continuous, quantile_bins, nbins, use_density, bin_method) %>%
    slice(1) %>%
    ungroup() %>%
    select(situation, bin_situation, bin_method, bin_continuous, quantile_bins, nbins, use_density, 
           set, case, rf_features, rfscore, model_prediction, model_r2) %>%
    mutate(diff = rfscore - model_prediction,
           mean = (rfscore + model_prediction) / 2)
